---
title: "biostats776-Project2-RMarkdown version"
author: "Zhengting (Johnathan) He"
date: "2021/9/27"
output: html_document
---

```{r "setup", include = FALSE}
require("knitr")
opts_knit$set(root.dir = "D:/OneDrive - Johns Hopkins/Course/140.776.71 - Statistical Computing/project/Project 2/biostats776-Project2")
```

```{r, message = FALSE}
# Set working directory and load packages
setwd("D:/OneDrive - Johns Hopkins/Course/140.776.71 - Statistical Computing/project/Project 2/biostats776-Project2")
require(tidyverse)
```


# Part 1: Fun with functions

In this part, we are going to practice creating functions.

### Part 1A: Exponential transformation 
The exponential of a number can be written as an infinite series expansion of the form
$$
\exp(x) = 1 + x + \frac{x^2}{2!} + \frac{x^3}{3!} + \cdots
$$
Of course, we cannot compute an infinite series by the end of this term and so we must truncate it at a certain point in the series. The truncated sum of terms represents an approximation to the true exponential, but the approximation may be usable.

Write a function that computes the exponential of a number using the truncated series expansion. The function should take two arguments:

* `x`: the number to be exponentiated

* `k`: the number of terms to be used in the series expansion beyond the constant 1. The value of `k` is always $\geq 1$.

For example, if $k = 1$, then the `Exp` function should return the number $1 + x$. If $k = 2$, then you should return the number $1 + x + x^2/2!$.

* You can assume that the input value `x` will always be a *single* number. 

* You can assume that the value `k` will always be an integer $\geq 1$. 

* Do not use the `exp()` function in R.

* The `factorial()` function can be used to compute factorials.


```{r}
Exp <- function(x, k) {
           sum(sapply(c(0:k), function(t) x^t/factorial(t)))
}
```


### Part 1B: Sample mean and sample standard deviation

Next, write two functions called `sample_mean()` and `sample_sd()` that takes as input a vector of data of length $N$ and calculates the sample average and sample standard deviation for the set of $N$ observations. 

$$
\bar{x} = \frac{1}{N} \sum_{i=1}^n x_i
$$
$$
s = \sqrt{\frac{1}{N-1} \sum_{i=1}^N (x_i - \overline{x})^2}
$$

* You can assume that the input value `x` will always be a *vector* of numbers of length $N$. 

* Do not use the `mean()` and `sd()` functions in R.

```{r}
sample_mean <- function(x) {
                   sum(x)/length(x)
}
sample_sd <- function(x) {
                 sqrt(sum(sapply(x, function(t) (t - sum(x)/length(x))^2))/(length(x) - 1))
}
```


### Part 1C: Confidence intervals

Next, write a function called `calculate_CI()` that:

1. Has two inputs to the `calculate_CI()` function. First, it should take as input a vector of data of length $N$. Second, the function should also have a `conf` ($=1-\alpha$) argument that allows the confidence interval to be adapted for different $\alpha$.


2. Calculates a confidence interval (CI) (e.g. a 95% CI) for the estimate of the mean in the population.


3. Returns a named vector of length 2, where the first value is the `lower_bound`, the second value is the `upper_bound`.


```{r}
calculate_CI <- function(x, conf) {
                    mean <- sample_mean(x)
                    sd_mean <- sample_sd(x)/sqrt(length(x))
                    t_score = qt(p = (1 - conf)/2, df = length(x) - 1, lower.tail = FALSE)
                    output <- matrix(c(mean - t_score*sd_mean, mean + t_score*sd_mean),
                                     nrow = 1,
                                     ncol = 2,
                                     dimnames = list("(Intercept)",
                                                     c("2.5 %", "97.5 %")))
                    return(output)
}
```


Checking code:


```{r}
x <- c(rnorm(100))

print("# Using my function `calculate_CI`")
calculate_CI(x, conf = 0.95)

print("# Using `confint` function in {base}")
dat = data.frame(x=x)
fit <- lm(x ~ 1, dat)
confint(fit, level=0.95)
```




# Part 2: Wrangling data

In this part, we will practice our wrangling skills with the tidyverse that we learned about in module 1. 

### Data 

The two datasets for this part of the assignment comes from [TidyTuesday](https://www.tidytuesday.com). 
Specifically, we will use the following data from January 2020, which I have provided for you below:  

```{r, eval=FALSE}
tuesdata <- tidytuesdayR::tt_load('2020-01-07')
rainfall <- tuesdata$rainfall
temperature <- tuesdata$temperature
```

**Note**: A useful way to avoid re-downloading data is to write code to check to see if those files already exist using and `if()` statement: 

```{r, message=FALSE}
if(!file.exists(("./data/tuesdata_rainfall.RDS"))){
    tuesdata <- tidytuesdayR::tt_load('2020-01-07')
    rainfall <- tuesdata$rainfall
    temperature <- tuesdata$temperature
  
    # save the files to RDS objects
    saveRDS(tuesdata$rainfall, file = "./data/tuesdata_rainfall.RDS")
    saveRDS(tuesdata$temperature, file = "./data/tuesdata_temperature.RDS")
}
```

**Note** the above code will only run if it cannot find the path to the `tuesdata_rainfall.RDS` on your computer. Then, we can just read in these files every time we knit the R Markdown, instead of redownloading them every time. 

```{r, eval=TRUE, message=FALSE}
rainfall <- readRDS("./data/tuesdata_rainfall.RDS")
temperature <- readRDS("./data/tuesdata_temperature.RDS")
```

Now we can look at the data with `glimpse()`

```{r,message=FALSE}
require(tidyverse)
glimpse(rainfall)
glimpse(temperature)
```

If we look at the [TidyTuesday github repo](https://github.com/rfordatascience/tidytuesday/tree/master/data/2020#2020-data) from 2020, we see this dataset contains temperature and rainfall data from Australia.

Here is a data dictionary for what all the column names mean: 

- https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-01-07/readme.md#data-dictionary

### Tasks 

Using the `rainfall` and `temperature` data, perform the following steps and create a new data frame called `df`: 

1. Start with `rainfall` dataset and drop any rows with NAs. 
2. Create a new column titled `date` that combines the columns `year`, `month`, `day` into one column separated by "-". (e.g. "2020-01-01"). This column should not be a character, but should be recognized as a date. (**Hint** check out the `ymd()` function in `lubridate` R package). You will also want to add a column that just keeps the `year`. 
3. Using the `city_name` column, convert the city names (character strings) to all upper case. 
4. Join this wrangled rainfall dataset with the `temperature` dataset such that it includes only observations that are in both data frames. (**Hint** there are two keys that you will need to join the two datasets together). (**Hint**: If all has gone well thus far, you should have a dataset with 83,964 rows and 13 columns). 

```{r}
df <- rainfall %>%
          drop_na() %>%
          unite(col = date, year, month, day, sep = "-") %>%
          mutate(date = lubridate::ymd(date),
                 year = lubridate::year(date)) %>%
          mutate(city_name = str_to_upper(city_name)) %>%
          inner_join(temperature, by = c("city_name", "date"))
glimpse(df)
head(df)
```


### Notes

* You may need to use functions outside these packages to obtain this result, in particular you may find the functions `drop_na()` from `tidyr` and `str_to_upper()` function from `stringr` useful.

 You may need to use functions outside these packages to obtain this result, in particular you may find the functions `drop_na()` from `tidyr` and `str_to_upper()` function from `stringr` useful.


# Part 3: Data visualization

In this part, we will practice our `ggplot2` plotting skills within the tidyverse that we learned about in module 1 starting with our wrangled `df` data from Part 2. For full credit in this part (and for all plots that you make), your plots should include: 

1. An overall title for the plot and a subtitle summarizing key trends that you found. Also include a caption in the figure. 
2. There should be an informative x-axis and y-axis label.

Consider playing around with the `theme()` function to make the figure shine, including playing with background colors, font, etc.

### Part 3A: Plotting temperature data over time

Use the functions in `ggplot2` package to make a line plot of the max and min temperature (y-axis) over time (x-axis) for each city in our wrangled data from Part 2. You should only consider years 2014 and onwards. For full credit, your plot should include: 

1. For a given city, the min and max temperature should both appear on the plot, but they should be two different colors. 
2. Use a facet function to facet by `city_name` to show all cities in one figure.

```{r}
df %>%
    filter(year >= 2014) %>%
    ggplot(aes(x = date, y = temperature, color = temp_type)) +
    geom_line() +
    facet_grid(city_name ~ ., scales = "free") +
    scale_color_manual("Temperature Type (Daily)",
                       values = c("#bb86fc", "#03dac5"),
                       labels = c("Max", "Min")) +
    labs(x = "Date",
         y = "Temperature in Celsius",
         title = "Daily Temperature in year 2014 and onwards in Australia",
         subtitle = "The Daily Temperature Fluctuate Between Each Season In Each Year, \nWith Highest Value in Winter and Lowest Value in Summer",
         caption = "By Zhengting (Johnathan) He") +
    theme(plot.background = element_rect(fill = "#262525", colour = NA),
          panel.background = element_rect(fill = "#121212", colour = NA),
          panel.border = element_rect(fill = NA, colour = "#1f1b24"),
          panel.grid = element_line(colour = "#1f1b24"), 
          panel.grid.minor = element_line(size = rel(0.5)),
          strip.background = element_rect(fill = "#000000", colour = "grey20"),
          legend.background = element_rect(fill = "transparent", colour = "transparent"),
          legend.key = element_rect(fill = "transparent"),
          plot.title = element_text(size = 12, face = "bold", color = "#ffffff"),
          plot.subtitle = element_text(size = 10, face = "bold", color = "#ffffff"),
          plot.caption = element_text(size = 10, color = "#ffffff"),
          axis.title.x = element_text(size = 10, face = "bold", color = "#ffffff"),
          axis.title.y = element_text(size = 10, face = "bold", color = "#ffffff"),
          axis.text.x = element_text(size = 8, color = "#ffffff"),
          axis.text.y = element_text(size = 8, color = "#ffffff"),
          strip.text.y = element_text(size = 8, color = "#ffffff", angle = 0),
          legend.position = "top",
          legend.title = element_text(size = 10, face = "bold", color = "#ffffff"),
          legend.text = element_text(size = 10, face = "bold", color = "#ffffff"))
```


### Part 3B: Plotting rainfall over time

Here we want to explore the distribution of rainfall (log scale) with histograms for a given city (indicated by the `city_name` column) for a given year (indicated by the `year` column) so we can make some exploratory plots of the data. **Note**: you are again using the wrangled data from Part 2.

The following code plots the data from one city (`city_name == "PERTH"`) in a given year (`year == 2000`). 

```{r}
df %>% 
  filter(city_name == "PERTH", year == 2000) %>% 
  ggplot(aes(log(rainfall))) + 
    geom_histogram()
```

While this code is useful, it only provides us information on one city in one year. We could cut and paste this code to look at other cities/years, but that can be error prone and just plain messy.

The aim here is to **design** and **implement** a function that can be re-used to visualize all of the data in this dataset. 

1. There are 2 aspects that may vary in the dataset: The **city_name** and the **year**. Note that not all combinations of `city_name` and `year` have measurements.

2. Your function should take as input two arguments **city_name** and **year**.

3. Given the input from the user, your function should return a **single** histogram for that input. Furthermore, the data should be **readable** on that plot so that it is in fact useful. It should be possible visualize the entire dataset with your function (through repeated calls to your function).

3. If the user enters an input that does not exist in the dataset, your function should catch that and report an error (via the `stop()` function).

For your homework submission

1. Write a short description of how you chose to design your function and why. 

I chose to design the function by:

+ Transform the user input to the object class needed, i.e., city_name to `character`, and year to `numeric`.

+ Check whether the user inputted city name and year meets with any possible observation in `df`.

  - I also figure out another way by generating all possible combinations of city_name and year (like what I did in Part 3B 3.), and to combine the user inputted city name and year to a vector and test whether it matches any possible combination, but I think directly using `filter(df, city_name == user.city.name & year == user.year)` and check whether it retrieve any observation is better.
  
+ If the user input meets with possible obs in `df`, plot the histogram. I change the `xlab`, `ylab`, `title` to try to make it readable.

+ If the user input does not meet with any obs in `df`, stop the function and print out an error message.

+ Test the function by generating a random city name and year that match an observation of `df`, and by intentionally inputting a wrong city name and year.

2. Present the code for your function in the R markdown document.

```{r}
city.year.hist <- function(user.city.name, user.year) {
    # Transform user input
    user.city.name <- as.character(user.city.name)
    user.year <- as.numeric(user.year)
    
    # Check whether user input retrieve existing combination of "city_name" and "year"
    # Plot histogram if yes, stop if not
    if(nrow(filter(df, city_name == user.city.name & year == user.year)) > 0) {
           return(
               df %>%
                   filter(city_name == user.city.name, year == user.year) %>%
                   ggplot(aes(log(rainfall))) +
                   geom_histogram(fill = "#bb86fc", colour = "#d5b3fe") +
                   labs(x = "Rainfall in Millimeters On Log Scale",
                        y = "Count of Days",
                        title = paste("Histogram of Yearly Rainfall for",
                                      user.city.name, "in", user.year, "in Australia",
                                      sep = " "),
                        subtitle = "The Distribution of the Frequency Scattered Between Rainfall On Log Scale",
                        caption = "By Zhengting (Johnathan) He") +
                   theme(plot.background = element_rect(fill = "#262525", colour = NA),
                         panel.background = element_rect(fill = "#121212", colour = NA),
                         panel.border = element_rect(fill = NA, colour = "#1f1b24"),
                         panel.grid = element_line(colour = "#1f1b24"), 
                         panel.grid.minor = element_line(size = rel(0.5)),
                         plot.title = element_text(size = 12, face = "bold", color = "#ffffff"),
                         plot.subtitle = element_text(size = 10, face = "bold", color = "#ffffff"),
                         plot.caption = element_text(size = 10, color = "#ffffff"),
                         axis.title.x = element_text(size = 10, face = "bold", color = "#ffffff"),
                         axis.title.y = element_text(size = 10, face = "bold", color = "#ffffff"),
                         axis.text.x = element_text(size = 8, color = "#ffffff"),
                         axis.text.y = element_text(size = 8, color = "#ffffff"))
           )
    } else stop("Invalid City Name or Year, Please Recheck Your Input")
}
```

3. Include at least one example of output from your function. 

```{r}
# Generate random city and year for testing
city_year_list <- df %>%
                      select(city_name, year) %>%
                      distinct() %>%
                      unite(col = city_year, city_name, year, sep = " and ") %>%
                      pull(city_year)
city_year_list <- str_split_fixed(city_year_list, " and ", 2)

# Test 1: correct
user.city.year <- city_year_list[sample(nrow(city_year_list), 1),]
print(user.city.year)
city.year.hist(user.city.year[1], user.city.year[2])
```

```{r error = TRUE}
# Test 2: error
city.year.hist("Baltimore", 2020)
```


# Part 4: Apply functions and plot

### Part 4A: Tasks

In this part, we will apply the functions we wrote in Part 1 to our rainfall data starting with our wrangled `df` data from Part 2.

1. First, filter for only years including 2014 and onwards.
2. For a given city and for a given year, calculate the sample mean (using your function `sample_mean()`), the sample standard deviation (using your function `sample_sd()`), and a 95% confidence interval for the average rainfall (using your function `calculate_CI()`). Specifically, you should add two columns in this summarized dataset: a column titled `lower_bound` and a column titled `upper_bound` containing the lower and upper bounds for you CI that you calculated (using your function `calculate_CI()`). 
3. Call this summarized dataset `rain_df`. 

```{r}
rain_df <- df %>%
               filter(year >= 2014) %>%
               group_by(city_name, year) %>%
               summarise(mean = sample_mean(rainfall),
                         sd = sample_sd(rainfall),
                         lower_bound = calculate_CI(rainfall, 0.95)[1],
                         upper_bound = calculate_CI(rainfall, 0.95)[2])
rain_df %>% knitr::kable()
```


### Part 4B: Tasks

Using the `rain_df`, plots the estimates of mean rainfall and the 95% confidence intervals on the same plot. There should be a separate faceted plot for each city. Think about using `ggplot()` with both `geom_point()` (and `geom_line()` to connect the points) for the means and `geom_errorbar()` for the lower and upper bounds of the confidence interval. 

```{r}
rain_df %>%
    ggplot(aes(x = year, color = city_name)) +
    geom_point(aes(y = mean), size = 2) +
    geom_line(aes(y = mean), size = 1) +
    geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound, width = 0.1)) +
    facet_wrap(~ city_name, scales = "free") +
    labs(x = "Year",
         y = "Summary Statistics of Rainfall (In Millimeters)",
         title = "Annual Rainfall in 5 Cities in Australia (2014-2019)",
         caption = "By Zhengting (Johnathan) He") +
    scale_color_manual("City",
                      values = c("#314855", "#5cc3e8", "#ffdb00", "#79ceb8", "#e95f5c")) +
    theme(panel.background = element_rect(fill='transparent', color='black'),
          panel.border = element_rect(fill='transparent', color='grey20'),
          panel.grid = element_line(colour = "grey92"), 
          panel.grid.minor = element_line(size = rel(0.5)),
          legend.background = element_rect(fill = "transparent", colour = "transparent"),
          legend.key = element_rect(fill = "transparent"),
          axis.title.x = element_text(size = 8, face = "bold", color = "black"),
          axis.title.y = element_text(size = 8, face = "bold", color = "black"),
          axis.text = element_text(size = 8, face = "bold", color = "black"),
          legend.title = element_text(size = 10, face = "bold", color = "black"),
          legend.text = element_text(size = 10, face = "bold", color = "black"),
          legend.position = "bottom",
          plot.caption = element_text(size = 8, face = "bold", color = "black"),
          strip.background = element_rect(fill='transparent', color='black'),
          strip.text.x = element_text(size = 8, face = "bold", color = "black"))
```
